Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews

### 2025-04-18 README.Rmd
revisions after peer review

1. Install [Docker](https://www.docker.com/get-started/) on your computer 
2. Download the compendium, e.g. from GitHub by clicking on the green `< > Code` button above, then click `Download ZIP`, then unzip on your computer
3. Set your terminal working directory to the compendium, and run `docker build -t wos .` to build the container on your computer, it will take a few minutes and require an internet connection.
4. Run in your terminal `docker run --rm -it -e ROOT=TRUE -e PASSWORD=rstudio -dp 8787:8787 wos` to start the container on your computer
5. Go to http://localhost:8787/ with your browser and enter `rstudio` in both fields to start RStudio in your browser, running in the Docker container with all the required depedencies installed
6. In RStudio, run `rstudioapi::openProject("/home/rstudio/web-of-science-archaeology")` to open the compendium's project
7. Run `rmarkdown::render("analysis/paper/paper.qmd")` to render the Quarto document that is the manuscript submitted for publication
8. When finished, to delete the Docker container from your computer, run this in the terminal `docker ps -aq | xargs docker stop | xargs docker rm`

>>>>

1. Install [Docker](https://www.docker.com/get-started/) on your computer 
2. Download the compendium, e.g. from GitHub by clicking on the green `< > Code` button above, then click `Download ZIP`, then unzip on your computer
3. Set your terminal working directory to the compendium, and run `docker build -t wos .` to build the container on your computer, it will take a few minutes and require an internet connection.
4. Run in your terminal `docker run --rm -it -e ROOT=TRUE -e PASSWORD=rstudio -dp 8787:8787 wos` to start the container on your computer
5. Go to http://localhost:8787/ with your browser and enter `rstudio` in both fields to start RStudio in your browser, running in the Docker container with all the required depedencies installed
6. In RStudio, run `rstudioapi::openProject("/home/rstudio/web-of-science-archaeology")` to open the compendium's project
7. Run ` rmarkdown::render("analysis/paper/paper.qmd", output_dir='/tmp')` to render the Quarto document that is the manuscript submitted for publication
8. After the code has executed, view the rendered output document by running `browseURL("/tmp/paper.html")`
8. When finished, to delete the Docker container from your computer, run this in the terminal `docker ps -aq | xargs docker stop | xargs docker rm`

### 2025-04-18 analysis/paper/paper.qmd
revisions after peer review

To objectively quantify the relative hardness or softness of archaeology, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach. This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters. @fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources. In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, less hard sciences include social sciences include Psychiatry, Psychology, Economics, Business and General Social Sciences, and the Humanities define the soft end of the spectrum. Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.

These parameters are useful because of how they signify consensus in a research community. A larger number of authors on a paper reflects collaboration of people working together on a common goal. Collaborators have specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail, with harder fields having larger groups of collaborators [@zuckerman1972age]. Reflecting this collaboration group size, harder disciplines tend to have higher average numbers of authors on papers. Article length has an inverse correlation with field hardness. In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results. The number of substantive and informative words in an article's title tends to be positively correlated with article length in harder disciplines [@yitzhakiRelationTitleLength2002; @yitzhakiVariationInformativityTitles1997], reflecting a focus on empiricism and efficiency that is characteristic of high-consensus disciplines. While @yitzhakiRelationTitleLength2002 removed stop-words (e.g. prepositions, articles, conjunctions, etc.) to calculate article length, in order to generate results for comparison with @fanelliBibliometricEvidenceHierarchy2013 I follow his method of dividing the total word count of the article title by the total number of pages of the article to compute relative title length.

The age of works cited has long been used as a measure of a field's hardness [@moed1998new; @bornerAtlasScience2010], based on the assumption that harder fields assimilate new results more rapidly that softer fields [@price1970citation]. I calculated a recency of references index for each article (also known as the Price index), which is the proportion of all cited works that were published in the five years preceding the paper. The diversity of references

>>>>

To objectively quantify the diversity of modern archaeological practice across a scale of relative hardness or softness, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach. This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters. @fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources. In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, less hard sciences include social sciences include Psychiatry, Psychology, Economics, Business and General Social Sciences, and the Humanities define the soft end of the spectrum. Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.

These parameters are useful because of how they signify consensus in a research community. A larger number of authors on a paper reflects collaboration of people working together on a common goal. Collaborators have specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail, with harder fields having larger groups of collaborators [@zuckerman1972age]. Reflecting this collaboration group size, harder disciplines tend to have higher average numbers of authors on papers. Article length has an inverse correlation with field hardness. In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results. While article length is constrained by journal requirements, leaving individual authors with little freedom to vary, journal requirements are typically set by editors who are professional archaeologists keen to tailor their journal to be attractive and relevant other members of the discipline. Thus journal requirements for article length will reflect the norms and of the discipline at any given time. The number of substantive and informative words in an article's title tends to be positively correlated with article length in harder disciplines [@yitzhakiRelationTitleLength2002; @yitzhakiVariationInformativityTitles1997], reflecting a focus on empiricism and efficiency that is characteristic of high-consensus disciplines. While @yitzhakiRelationTitleLength2002 removed stop-words (e.g. prepositions, articles, conjunctions, etc.) to calculate article length, in order to generate results for comparison with @fanelliBibliometricEvidenceHierarchy2013 I follow his method of dividing the total word count of the article title by the total number of pages of the article to compute relative title length.

The age of works cited has long been used as a measure of a field's hardness [@moed1998new; @bornerAtlasScience2010], based on the assumption that harder fields assimilate new results more rapidly that softer fields [@price1970citation]. I calculated a recency of references index for each article (also known as the Price index), which is the proportion of all cited works that were published in the five years preceding the paper. The diversity of references is a similar indicator, with papers in harder fields having a higher concentration of more specific citations because more knowledge is taken for granted as core knowledge that does not need citing [@skilton2006comparative]. Conversely, softer fields have less knowledge taken for granted, a smaller core of facts that do not need citing, and thus a higher diversity of citations.
